---
title: "D2"
output:
  pdf_document: default
  html_document: default
---
# (Unsupervised learning) clustering

## The hierarchical clustering

this method is implemented in R within the 'class' package and the appropriate method is names 'hclust'

>exercise: cluster the 'swiss' data with 'hclust'

```{r}
library(class)
x=swiss
#?hclust
#need to build distance matrixs 
dx=dist(x) #default: method"euclidean"

#ward distance
cluster<-hclust(dx,method="ward.D2")
#clearly 2 groups
plot(cluster)
```
now try 4 differenct method
```{r}
par(mfrow=c(2,2))
#single distance
cluster1<-hclust(dx,method="single")
plot(cluster1)
#complete distance
outComplete<-hclust(dx,method="complete")
plot(outComplete)
#centroiddistance
cluster3<-hclust(dx,method="centroid")
plot(cluster3)
#ward distance
outWard<-hclust(dx,method="ward.D2")
plot(outWard)
```



* now we choose only 2 to compare: complete and ward
at this point we don't have yet the assignment to the clustering: we need cutree
```{r}
plot(outComplete)
k1=3
#get clustering
res1=cutree(outComplete,k1)
#visualize for cluster
rect.hclust(outComplete,k1)

plot(outWard)
k2=2
res2=cutree(outWard,k2)
rect2=rect.hclust(outWard,k2)

```


make a pair to see the variable result compaire to the clustering
```{r}
pairs(swiss,col=res1,pch=19)
pairs(swiss,col=res2,pch=19)
```

## The mixture model and the EM algorithm

the 'mclust' package (Raftery et al.) allow to cluster some data with GGM and the EM algorithm
```{r}
#install.packages('mclust')
library(mclust)
```

```{r}
data("swiss")
#G=number of group
out=Mclust(swiss, G=2:10)
plot(out)

# plot 1:
#get the highest point of BIC (3 groups)
#here best model is EEE, which is exactly K-mean


# plot 2:
# we see that the result is exactly the same as HC (complete)

#plot 3:
# the larger is the point the larger is the uncertainty

```

```{r}
out$modelName
out$parameters$mean
```

```{r}
out$parameters$pro
out$parameters$variance
```

the Rmix package also allows to use the GGM+EM

```{r}
#install.packages('Rmixmod')
library(Rmixmod)
```
```{r}
out=mixmodCluster(swiss,2:10)
# 2:10 = means that it would choose the best group between it
#default is 1 to 9. if it's 1 it means that there's no need to do clustering
#plot(out) # type in the console
```

