---
title: "D1 K Mean"
output:
  html_document: default
  pdf_document: default
---
# reminder about the descriptive statisic
```{r}
x=iris
summary(x)
```

As we saw, descriptive statistics are useful to start discovering the data
(here is obvious a supervise learning)

* about histogram: best choice by (bins number)
```{r}
hist(x$Sepal.Length)
```

#best choice by ggplot (bins number)
```{r}
library(ggplot2)
ggplot(x)+geom_histogram(aes(x=Sepal.Length))
```

```{r}
ggplot(x)+geom_histogram(aes(x=Sepal.Length),bins=100)
hist(x$Sepal.Length,breaks=20,freq=FALSE)
lines(density(x$Sepal.Length),col='red')
```

* boxplot
check petal.length, here we have a distribution, very low to increase and very fast to decrease
```{r}
barplot(summary(x$Species))
```
* try to pair to see the relationship
```{r}
pairs(x[,-5],col=as.numeric(x$Species))
```


# Cluster
## K-means

The K means algorithm is provided in the 'class package ' and the function is named 'kmeans'
```{r}
library(class)
?kmeans
```

```{r}
#try 1. not set nstart, 2. try nstart=10 not a good result
out=kmeans(x[,-5],3,nstart=10)
out
```

let's try to find the most appropriate number of groups:
```{r}
J=c()
for (k in 1:15){
  out=kmeans(x[,-5],k,nstart=15)
  J[k]=out$betweenss/out$totss   #B/S
}
plot(1:15,J,type='b') 
#we choose k from 2-15, so
```

here we shoud choose wither 3 or 4 groups
```{r}
out= kmeans(x[,-5],4)
pairs(x[,-5],col=out$cluster,pch=19)
out1= kmeans(x[,-5],3)
pairs(x[,-5],col=out1$cluster,pch=19)
```

> exercuse: use the k-means to cluster 'swiss' data

```{r}
x1=swiss
summary(swiss)
```
```{r}
J1=c()
for (k in 1:15){
  out=kmeans(x1,k,nstart=15)
  J1[k]=out$betweenss/out$totss   #B/S
}
plot(1:15,J,type='b') 
```

I assume either 4 or 5 groups to choose
```{r}
out11= kmeans(x1,4,nstart=15)
pairs(x1,col=out11$cluster,pch=19)
out12= kmeans(x1,5, nstart=15)
pairs(x1,col=out12$cluster,pch=19)

out12
#here we see that the cluster 2 means we think it is a big city, more balance
#there's geneve, Rive Droite, Rive Gauche
```



(short insert cut ctrl+alt+I)