---
title: "Supervised learning, Learnin in high dimention"

output: pdf_document
---
# Supervised learning

## The general procedure

let's condider some classification problem with the iris data

```{r}
library(MASS);library(class); library(e1071)
x=iris[,-5]
y=iris[,5]
E=matrix(NA,3,nrow(x))
# I would like to compare several classification technique (LDA, KNN, SVM)
#use leave one out method
for (b in 1:nrow(x)){
  xl=x[-b,];yl=y[-b]
  xv=x[b,];yv=y[b] #leave one out
  #LDA
  y1=predict(lda(xl,yl),xv)
  #KNN
  y2=knn(xl,xv,yl)
  #SVM
  y3=predict(svm(xl,yl),xv)
  
  E[1,b]= as.numeric(y1$class != yv)
  E[2,b]= as.numeric(y2 != yv)
  E[3,b]= as.numeric(y3 != yv)
}

#pick the most appropriate method
boxplot(t(E))

apply(E,1,mean)
apply(E,1,sd)
#smallest mean and smallest sd, lda is the best method

#my final pick lda
clf=lda(x,y)

#in the future when i have new observation
xstar=c(0.5, 8,56,4)
predict(clf,xstar)
```

# Learning in high-dimensional spaces

## The curs of dimensionality

The volume of the hyper-sphere is $V(p) = \frac{\pi^{p/2}}{\Gamma(p/2+1)}$.

```{r}
p = 1:100
plot(p,pi^(p/2) / gamma(p/2+1), type='b')
```

## High-dimensional data clustering (HDDC)

HDDC is implemented in the `HDclassif` library (which also p^roposes the HDDA method for classification).

```{r}
#install.packages('HDclassif')
library(HDclassif)
#?hddc
```

We may try HHDC on the wine data set:

```{r}
data(wine)
X = scale(wine[,-1])
Y = wine$class

out = hddc(X,K=1:10)#,threshold = (1:10)/100)
plot(out)
out$d
table(out$cl,Y)
```

Comparison with Mclust

```{r}
library(mclust)
out1 = hddc(X,K=3)
table(out1$cl,Y)

out2 = Mclust(X,G=3)
table(out2$cl,Y)

```

Let's now move to higher dimensions:

```{r}
#install.packages('MBCbook')
library(MBCbook)
data("usps358")
X = usps358[,-1]
Y = usps358$cls

system.time(out1 <- hddc(X,K=3))
table(out1$cl,Y)

#system.time(out2 <- Mclust(X,G=3,modelNames = 'VVV'))
#table(out2$cl,Y)
```

```{r}
library(MBCbook)
library(HDclassif)
data("usps358")
X = usps358[,-1]
Y = usps358$cls

out1 = hddc(X,K=3)
table(out1$cl,Y)

image(matrix(out1$mu[1,],ncol=16))
image(matrix(out1$mu[2,],ncol=16))
image(matrix(out1$mu[3,],ncol=16))
```

Fisher EM
```{r}
#install.packages('FisherEM')
library(FisherEM)
out=fem(X,K=3)
table(out$cl,Y)
plot(out)

```



